---
title: "`R` Installation"
subtitle: "HORT / AGRO 4401"
date: "March 7, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)

# Load libraries
library(tidyverse)
library(png)
library(grid)

```

## Introduction

This guide will provide steps for installing R and the quantitative genetics demo.

## Installing R

1. Navigate to The R Project website: https://www.r-project.org/

```{r, fig.align='center'}

png_name <- "figures/step1.PNG"
img <- readPNG(png_name)
grid.raster(img)

```

\newpage

2. Click on "download R" (or follow this link: https://cran.r-project.org/mirrors.html)

```{r, fig.align='center'}

png_name <- "figures/step2.PNG"
img <- readPNG(png_name)
grid.raster(img)

```

\newpage

3. Choose a mirror for download. I usually use the Iowa State mirror (or follow this link: https://mirror.las.iastate.edu/CRAN/)

```{r, fig.align='center'}

png_name <- "figures/step3.PNG"
img <- readPNG(png_name)
grid.raster(img)

```

\newpage

4. Choose the download the is most appropriate for your operating system.

    + If using a Mac, click *Download R for (Mac) OS X*, download the **.pkg** file and install
  
```{r, fig.align='center'}

png_name <- "figures/step4_mac.PNG"
img <- readPNG(png_name)
grid.raster(img)

```

\newpage

    + If using a Windows machine, click *Download R for Windows*, and click *install R for the first time*. Then click the Download link at the top of the page and install.
    
```{r, fig.align='center'}

png_name <- "figures/step4_pc.PNG"
img <- readPNG(png_name)
grid.raster(img)

```


\newpage

## Installing Shiny

1. Once R is installed, open the program.

2. Type the following command and press `Enter`:

```
install.packages("shiny")
```


